---
title: "Midterm - Steve Cheney"
output: html_document
date: "2024-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data inports
"143" = journal_1_.pone.0000216.s007
"159" = journal_1_.pone.0000216.s008

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

```{r}
# S3 / 143
s3_url <- "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1789144/bin/pone.0000216.s007.xls"
s3_file <- "table_s3_143.xls"
curl::curl_download(s3_url, s3_file)
s3_table <- read_excel(s3_file, range = "A4:U147")


# S4 / 159
s4_url <- "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1789144/bin/pone.0000216.s008.xls"
s4_file <- "table_s3_143.xls"
curl::curl_download(s4_url, s4_file)
s4_table <- read_excel(s4_file, range = "A4:U163")

rename_first_three_columns <- function(data_frame) {
  # Rename the first three columns
  names(data_frame)[1:3] <- c("refseq_id", "uniprot_swissprot", "protein_name")
  
  # Return the modified data frame
  return(data_frame)
}

s3_table <- rename_first_three_columns(s3_table)
s4_table <- rename_first_three_columns(s4_table)
```

```{r}
# Calculate the means for s3
s3_means <- s3_table %>%
  reframe(
    chimp_mrna = apply(s3_table[,4:9], 1, mean, na.rm = TRUE),
    human_mrna = apply(s3_table[,10:15], 1, mean, na.rm = TRUE),
    chimp_protein = apply(s3_table[,16:18], 1, mean, na.rm = TRUE),
    human_protein = apply(s3_table[,19:21], 1, mean, na.rm = TRUE)
  )
# Add the means to the s3 table
s3_table <- s3_table %>%
  mutate(
    chimp_mrna_mean = apply(s3_table[,4:9], 1, mean),
    human_mrna_mean = apply(s3_table[,10:15], 1, mean),
    chimp_protein_mean = apply(s3_table[,16:18], 1, mean),
    human_protein_mean = apply(s3_table[,19:21], 1, mean)
  )
# Calculate the means for s4
s4_means <- s4_table %>%
  reframe(
    chimp_mrna = apply(s4_table[,4:9], 1, mean, na.rm = TRUE),
    human_mrna = apply(s4_table[,10:15], 1, mean, na.rm = TRUE),
    chimp_protein = apply(s4_table[,16:18], 1, mean, na.rm = TRUE),
    human_protein = apply(s4_table[,19:21], 1, mean, na.rm = TRUE)
  )
# Add the means to the s4 table
s4_table <- s4_table %>%
  mutate(
    chimp_mrna_mean = apply(s4_table[,4:9], 1, mean),
    human_mrna_mean = apply(s4_table[,10:15], 1, mean),
    chimp_protein_mean = apply(s4_table[,16:18], 1, mean),
    human_protein_mean = apply(s4_table[,19:21], 1, mean)
  )
```

## Question 1
```{r}
# Plotting function
plot_data <- function(data, title_suffix) {
  p1 <- ggplot(data, aes(x = human_mrna_mean, y = chimp_mrna_mean)) +
    geom_point() +
    labs(title = paste("mRNA for", title_suffix),
         x = "Human mRNA Mean",
         y = "Chimp mRNA Mean")
  
  p2 <- ggplot(data, aes(x = human_protein_mean, y = chimp_protein_mean)) +
    geom_point() +
    labs(title = paste("Protein for", title_suffix),
         x = "Human Protein Mean",
         y = "Chimp Protein Mean")
  
  # Return plots as a list for grid
  list(mRNA = p1, Protein = p2)
}

# Generate plots for s3_table and s4_table
s3_plots <- plot_data(s3_table, "Sample S3 (143)")
s4_plots <- plot_data(s4_table, "Sample S4 (159)")

grid.arrange(s3_plots$mRNA, s3_plots$Protein, s4_plots$mRNA, s4_plots$Protein, ncol = 2)
```

In these plots you can see that there is a relatively strong correlation between the mRNA means in each sample for Human vs. Chimp and each hold to about the same overall means on the graph. There is less of a strong correlation between proteins, comparatively, and we start to potentially see the difference in the proteome where S3 has, in general, greater values for the Chimp Protein means in S3 compared to the S4 Chimp Protein means.

## Question 2
```{r}
# Calculate correlations with cor()
calculate_correlations <- function(data, sample_name) {
  # Calculating correlation between human and chimp mRNA means
  cor_mRNA <- cor(data$human_mrna_mean, data$chimp_mrna_mean, use = "complete.obs")
  
  # Calculating correlation between human and chimp protein means
  cor_protein <- cor(data$human_protein_mean, data$chimp_protein_mean, use = "complete.obs")
  
  cat(paste("\nCorrelation of mRNA between humans and chimps for", sample_name, ":", cor_mRNA))
  cat(paste("\nCorrelation of protein between humans and chimps for", sample_name, ":", cor_protein))
}

# Calculate correlations for s3_table
calculate_correlations(s3_table, "Sample S3")

# Calculate correlations for s4_table
calculate_correlations(s4_table, "Sample S4")

```

We can see here that I was correct in saying that both the mRNA sample comparisons had a high correlation of about 0.957 each. I was also correct in saying that the protein sample comparisons had a low correlation (and both negative) where S3 is about -0.5 and S4 is about -0.221. S4's protein means has the least amount of correlation among all sample comparisons.
